---
title: "Analysis of the age variable"
date: "2023"
author: "José Carlos Soares Junior"
geometry: "left=1cm,right=1cm,top=2cm,bottom=2cm"
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H} 
  - \renewcommand{\contentsname}{Contents}
  - \usepackage{mdframed}
  - \definecolor{shadecolor}{gray}{.90}
  - \renewenvironment{Shaded}{\begin{mdframed}[
      backgroundcolor=shadecolor,
      linecolor = shadecolor,
      leftmargin=\dimexpr\leftmargin-2pt\relax,
      innerleftmargin=1.6pt,
      innertopmargin=5pt,
      skipabove=10pt,skipbelow=3pt
    ]}{\end{mdframed}}
output:   
  pdf_document:
    toc: yes
    toc_depth: '3'
    keep_tex: yes
  word_document: default
  html_document:
    self_contained: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# R packages

```{r packages,echo=TRUE,message=FALSE,results='hide',warning=FALSE}
# Load packages
loadlibrary <- function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = T)
  if (!require(x, character.only = TRUE))
  stop("Package not found")
  }
}

packages <- c(
  "readr",
  "readxl",
  "janitor",
  "dplyr",
  "forcats",
  "stringr",
  "lubridate",
  "summarytools",
  "magrittr",
  "questionr",
  "knitr",
  "data.table",
  "writexl",
  "modelsummary",
  "abjData",
  "aweek",
  "ggplot2",
  "viridis"
)
lapply(packages, loadlibrary)

# Global changes for the summarytools packages
st_options(freq.cumul = FALSE,          
           display.labels = TRUE,
           lang = "pt",
           bootstrap.css = FALSE,
           plain.ascii = FALSE,
           dfSummary.silent = TRUE,
           footnote = NA,
           subtitle.emphasis = FALSE,
           headings = FALSE)
st_css()
```

\newpage

# Data preparation

```{r,echo=TRUE,message=FALSE,results='hide',warning=FALSE}
###################### Loading the data #######################
data_age <- readRDS("data_paper_1000d.rds")
```

```{r,echo=TRUE,message=FALSE,results='hide',warning=FALSE}
###################### age variable #######################

######## First case ########
# <= 30 days (1 month);
# 31 days |- 180 days (between 1 and 6 months); 
# 181 days |- 365 days (between 6 and 12 months); 
# >= 366 days (more than 12 months).

data_age1 <- data_age %>%
  mutate(age_months_firstcase = case_when(
    idade_dias <= 30 ~ "1 month",
    idade_dias > 30 & idade_dias <= 180 ~ "between 1 and 6 months",
    idade_dias > 180 & idade_dias <= 365 ~ "between 6 and 12 months",
    idade_dias > 365 ~ "more than 12 months") 
    )

######## Second case ########
# <= 30 days (1 month); 
# 31 days |- 180 days (between 1 and 6 months); 
# 181 days |- 365 days (between 6 and 12 months); 
# 366 days |- 545 days (between 12 and 18 months);
# >= 546 days (more than 18 months).

data_age2 <- data_age1 %>%
  mutate(age_months_secondcase = case_when(
    idade_dias <= 30 ~ "1 month",
    idade_dias > 30 & idade_dias <= 180 ~ "between 1 and 6 months",
    idade_dias > 180 & idade_dias <= 365 ~ "between 6 and 12 months",
    idade_dias > 365 & idade_dias <= 545 ~ "between 12 and 18 months",
    idade_dias > 545 ~ "more than 18 months") 
    )
```

# Quantitative analysis

## Analysis of age(days)

```{r,echo=TRUE,message=FALSE,warning=FALSE}
ggplot(data_age2, aes(x = idade_dias)) + 
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 125,
                 colour = "black",
                 fill = "white") +
  geom_density(alpha = .2, fill = "#FF6666") +
  labs(x = "\nAge(days)",
       y = "Density\n",
       title = "Histogram of age(days)")
```

**Figure 1:** Histogram of age in unit of days (bars) and density (curve with red area).

```{r,echo=TRUE,message=FALSE,results='asis',warning=FALSE}
data_age2$variants <- factor(data_age2$variants, 
                             levels = c("original",
                                        "gamma",
                                        "delta",
                                        "omicron")
                             )

data_age2 %>% 
  ggplot(aes(x = variants,
             y = idade_dias,
             fill = variants)) +
  geom_boxplot() +
  xlab("\nVariants") +
  ylab("Age(Days)\n") +
  theme(legend.position = "none")
```

**Figure 2:** Boxplot of variants by age in a unit of days.

```{r,echo=TRUE,message=FALSE,results='asis',warning=FALSE}
# Summary measurement functions
media <- function(x) 
  mean(x, na.rm = TRUE)
mediana <- function(x)
  median(x, na.rm = TRUE)
DP <- function(x)
  sd(x, na.rm = TRUE)
minimo <- function(x)
  base::min(x, na.rm = TRUE)
maximo <- function(x) 
  base::max(x, na.rm = TRUE)
q25 <- function(x) 
  stats::quantile(x, p = 0.25, na.rm = TRUE)
q75 <- function(x)
  stats::quantile(x, p = 0.75, na.rm = TRUE)
IQR <- function(x) 
  round(q75(x) - q25(x), 2)
n <- function(x) 
  sum(!is.na(x))
```

\newpage

```{r,echo=TRUE,message=FALSE,results='asis',warning=FALSE}
datasummary(variants ~ idade_dias*(n + media + DP + mediana + minimo +
                                   maximo + q25 + q75 + IQR), 
            data = data_age2, output = 'markdown')
```

**Table 1:** Descriptive metrics of age (days) by variants.

## Analysis of age(months)

```{r,echo=TRUE,message=FALSE}
ggplot(data_age2, aes(x = age_month)) + 
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 24,
                 colour = "black",
                 fill = "white") +
  geom_density(alpha = .2, fill = "#FF6666") +
  labs(x = "\nAge(month)",
       y = "Density\n",
       title = "Histogram of age(month)")
```

**Figure 3:** Histogram of age in unit of months (bars) and density (curve with red area).

```{r,echo=TRUE,message=FALSE,results='asis',warning=FALSE}
data_age2 %>% 
  ggplot(aes(x = variants,
             y = age_month,
             fill = variants)) +
  geom_boxplot() +
  xlab("\nVariants") +
  ylab("Age(month)\n") +
  theme(legend.position = "none")
```

**Figure 4:** Boxplot of variants by age in a unit of months.

```{r,echo=TRUE,message=FALSE,results='asis',warning=FALSE}
datasummary(variants ~ age_month*(n + media + DP + mediana + minimo +
                                   maximo + q25 + q75 + IQR), 
            data = data_age2, output = 'markdown')
```

**Table 2:** Descriptive metrics of age (months) by variants.

\newpage

# Grouped age

## First case

```{r,echo=TRUE,message=FALSE,results='asis',warning=FALSE}
######## First case ########
data_age2$age_months_firstcase <- 
  as.factor(data_age2$age_months_firstcase)

summarytools::ctable(
        data_age2$age_months_firstcase,
        data_age2$variants,
        prop = "c",
        headings = st_options("headings"),
        display.labels = st_options("display.labels"),
        useNA = "ifany",
        OR = TRUE,
        chisq = TRUE
      )
```

**Table 3:** Age(months) versus variants, where age is grouped as: <= 30 days (1 month); 31 days |- 180 days (between 1 and 6 months); 181 days |- 365 days (between 6 and 12 months); >= 366 days (more than 12 months).

## Second case

```{r,echo=TRUE,message=FALSE,results='asis',warning=FALSE}
######## Second case ########
data_age2$age_months_secondcase <- 
  factor(data_age2$age_months_secondcase, 
         levels = c("1 month",
                    "between 1 and 6 months",
                    "between 6 and 12 months",
                    "between 12 and 18 months",
                    "more than 18 months")
         )

summarytools::ctable(
        data_age2$age_months_secondcase,
        data_age2$variants,
        prop = "c",
        headings = st_options("headings"),
        display.labels = st_options("display.labels"),
        useNA = "ifany",
        OR = TRUE,
        chisq = TRUE
      )
```

**Table 4:** Age(months) versus variants, where age is grouped as: <= 30 days (1 month); 31 days |- 180 days (between 1 and 6 months); 181 days |- 365 days (between 6 and 12 months); 366 days |- 545 days (between 12 and 18 months); >= 546 days (more than 18 months).

# Analysis by evolution(death = yes)

## Quantitative analysis

### Age(days)

```{r,echo=TRUE,message=FALSE}
data_age2_death <- filter(data_age2, death == "death")

ggplot(data_age2_death, aes(x = idade_dias)) + 
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 50,
                 colour = "black",
                 fill = "white") +
  geom_density(alpha = .2, fill = "#FF6666") +
  labs(x = "\nAge(days)",
       y = "Density\n",
       title = "Histogram of age(days)")
```

**Figure 5:** Histogram of age in unit of days (bars) and density (curve with red area). Here, only cases that evolved into death are considered.

```{r,echo=TRUE,message=FALSE,results='asis',warning=FALSE}
data_age2_death %>% 
  ggplot(aes(x = variants,
             y = age_month,
             fill = variants)) +
  geom_boxplot() +
  xlab("\nVariants") +
  ylab("Age(days)\n") +
  theme(legend.position = "none")
```

**Figure 6:** Boxplot of variants by age in a unit of days. Here, only cases that evolved into death are considered.

```{r,echo=TRUE,message=FALSE,results='asis',warning=FALSE}
datasummary(variants ~ idade_dias*(n + media + DP + mediana + minimo +
                                   maximo + q25 + q75 + IQR), 
            data = data_age2_death, output = 'markdown')
```

**Table 5:** Descriptive metrics of age (days) by variants. Here, only cases that evolved into death are considered.

### Age(months)

```{r,echo=TRUE,message=FALSE}
ggplot(data_age2_death, aes(x = age_month)) + 
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 24,
                 colour = "black",
                 fill = "white") +
  geom_density(alpha = .2, fill = "#FF6666") +
  labs(x = "\nAge(month)",
       y = "Density\n",
       title = "Histogram of age(month)")
```

**Figure 7:** Histogram of age in unit of months (bars) and density (curve with red area). Here, only cases that evolved into death are considered.

```{r,echo=TRUE,message=FALSE,results='asis',warning=FALSE}
data_age2_death %>% 
  ggplot(aes(x = variants,
             y = age_month,
             fill = variants)) +
  geom_boxplot() +
  xlab("\nVariants") +
  ylab("Age(month)\n") +
  theme(legend.position = "none")
```

**Figure 8:** Boxplot of variants by age in a unit of months. Here, only cases that evolved into death are considered.

```{r,echo=TRUE,message=FALSE,results='asis',warning=FALSE}
datasummary(variants ~ age_month*(n + media + DP + mediana + minimo +
                                   maximo + q25 + q75 + IQR), 
            data = data_age2_death, output = 'markdown')
```

**Table 6:** Descriptive metrics of age (months) by variants. Here, only cases that evolved into death are considered.

## Grouped age

### First case

```{r,echo=TRUE,message=FALSE,results='asis',warning=FALSE}
######## First case ########
summarytools::ctable(
        data_age2_death$age_months_firstcase,
        data_age2_death$variants,
        prop = "c",
        headings = st_options("headings"),
        display.labels = st_options("display.labels"),
        useNA = "ifany",
        OR = TRUE,
        chisq = TRUE
      )
```

**Table 7:** Age(months) versus variants, where age is grouped as: <= 30 days (1 month); 31 days |- 180 days (between 1 and 6 months); 181 days |- 365 days (between 6 and 12 months); >= 366 days (more than 12 months). Here, only cases that evolved into death are considered.

### Second case

```{r,echo=TRUE,message=FALSE,results='asis',warning=FALSE}
######## Second case ########
summarytools::ctable(
        data_age2_death$age_months_secondcase,
        data_age2_death$variants,
        prop = "c",
        headings = st_options("headings"),
        display.labels = st_options("display.labels"),
        useNA = "ifany",
        OR = TRUE,
        chisq = TRUE
      )
```

**Table 8:** Age(months) versus variants, where age is grouped as: <= 30 days (1 month); 31 days |- 180 days (between 1 and 6 months); 181 days |- 365 days (between 6 and 12 months); 366 days |- 545 days (between 12 and 18 months); >= 546 days (more than 18 months). Here, only cases that evolved into death are considered.
